---
title: "Human Origins Array West Eurasia Results"
author: "jhmarcus"
date: "2019-02-15"
output:
  workflowr::wflow_html:
    code_folding: show
---

## Imports

Lets import some needed packages 

```{r imports, warning=FALSE, message=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(RColorBrewer)
source("../code/viz.R")
```

## Human Origins West Eurasia (LD Pruned)

This is a subset of the Human Origins Array dataset with 770 sampled from across West Eurasia. I filtered out rare variants with global minor allele frequency less than 5%, and remove any variants with a missingness level greater than 1%. I then LD pruned the SNPs using standard parameters in `plink`, resulting in 139640 SNPs.

### Greedy

Lets first read the greedy `flashier` fit

```{r flash-greedy-ld}
flash_fit = readRDS("../output/flash_greedy/hoa_weurasia_ld/HumanOriginsPublic2068_weurasia_maf_geno_ldprune.rds")
K = ncol(flash_fit$loadings$normalized.loadings[[1]]) 
n = nrow(flash_fit$loadings$normalized.loadings[[1]])
p = nrow(flash_fit$loadings$normalized.loadings[[2]])
print(K)
print(n)
print(p)
```

Lets now plot the distribution of factors for each drift event

```{r flash-greedy-ld-viz-factors, warning=FALSE, message=FALSE, fig.width=8, fig.height=6}
# read factors
delta_df = as.data.frame(flash_fit$loadings$normalized.loadings[[2]])
colnames(delta_df)[1:K] = 1:K 

# gather the data.frame for plotting
delta_gath_df = delta_df %>% 
                gather(K, value) %>%
                filter(K!=1)

# plot the factors
K_ = K
p_fct = ggplot(delta_gath_df, aes(x=value)) + 
        scale_fill_manual(values = getPalette(K_)) +
        geom_histogram() + 
        facet_wrap(~factor(K, levels=2:K_), scales = "free") + 
        labs(fill="K") + 
        scale_x_continuous(breaks = scales::pretty_breaks(n = 3)) +
        scale_y_continuous(breaks = scales::pretty_breaks(n = 3)) + 
        theme_bw()
p_fct
```

We can see the later factors tend to get sparser but they still seem to contribute! Here is a plot of the "proportion of variance explained" of each factor:

```{r flash-greedy-ld-viz-pve}
qplot(2:K, flash_fit$pve[2:K]) + ylab("Proportion of Varaince Explained") + xlab("K") + theme_bw()
print(flash_fit$pve)
```

I setup the `flashier` run so it estimates a SNP speficic precision term. Here is a histogram of fitted variances:

```{r flash-greedy-ld-viz-var}
qplot(1/flash_fit$fit$est.tau) + xlab("Estiamted Variance") + ylab("Count") + theme_bw()
```

This looks reasonable as each of the SNP variances should rougly be interpreted as the average heterozygosity $\approx 2p(1-p)$. The mean term should roughly be interpreted as the mean allele frequeny at the SNP and thus we should see a quadratic relationship with the estimated variance:

```{r flash-greedy-ld-viz-mv}
p_mv = qplot(delta_df$`1`, 1/flash_fit$fit$est.tau, alpha=.3) + 
       xlab("Estimated Mean") + ylab("Estimated Variance") + 
       scale_alpha(guide = "none") + 
       theme_bw()
p_mv
```

Lets now take a look at the loadings. First we setup a data.frame that we can work with:

```{r flash-greedy-ld-data-loadings, warning=FALSE, message=FALSE}
# read the meta data
meta_df = read.table("../data/meta/HumanOriginsPublic2068_weurasia_maf_geno_ldprune.meta", sep=" ", header=T)

# setup loadings data.frame
l_df = as.data.frame(flash_fit$loadings$normalized.loadings[[1]])
l_df$iid = as.vector(meta_df$iid) # individual ids
l_df$clst = meta_df$clst # population labels

# join with the meta data
l_df = l_df %>% inner_join(meta_df, on="clst")
l_df = l_df %>% arrange(region, clst) # sort by region then by population
l_df$iid = factor(l_df$iid, levels = l_df$iid) # make sure the ids are sorted
colnames(l_df)[1:K] = 1:K

pops = unique(l_df$clst)

head(l_df)
```

```{r flash-greedy-ld-viz-loadings-2-10}
# gather the data.frame for plotting
l_gath_df = l_df %>% 
            gather(K, value, -iid, -clst, -region, -country, -lat, -lon, -clst2) %>% 
            filter(K %in% paste0(2:10))

pops = get_pops(meta_df, "WestEurasia")
p = positive_structure_plot(l_gath_df, pops, colset="Set3", label_size=5)
```