---
title: 'Branching + admixture simulation: large p and longer runs'
author: "Jason Willwerscheid"
date: "1/10/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 5, fig.height = 4)
```

```{r load.pkgs}
suppressMessages({
  library(flashier)
  library(drift.alpha)
  library(ggplot2)
  library(reshape2)
  library(tidyverse)
})
```

The setup is the same as the [previous simulation](admix_sim4.html), but I've increased $p$ to a more realistic 10000 and in each case I run `drift` for a minimum of 1000 iterations. I only include the more promising initializations.

```{r sim}
set.seed(666)
simple.admix <- admix_graph_sim(n_per_pop = 20, p = 10000, 
                                c1 = 2, c2 = 1, c3 = 0, c4 = 0,
                                c5 = 1, c6 = 1, c7 = 0,
                                w = 0.5, sigma_e = sqrt(0.25))
plot_cov(simple.admix$covmat, as.is = TRUE)
```

## Flash initialization (greedy)

### Initial values

```{r flg1}
fl <- flash(simple.admix$Y, prior.family = c(prior.bimodal(), prior.normal()))
labs <- rep(c("A", "B", "C", "D"), each = 20)
plot_loadings(fl$flash.fit$EF[[1]], labs)
```

### Drift results

```{r flg2}
drift.flg <- drift(init_from_flash(fl), miniter = 1000, maxiter = 1000, tol = 1e-4, verbose = FALSE)
ggplot(drift.flg$iterations, aes(x = iter, y = elbo)) + geom_line()
drift.flg[c("elbo", "KL_l", "KL_f")]

plot_loadings(drift.flg$EL, labs,  paste("s2:", round(drift.flg$prior_s2, 2)))
plot_cov(drift.flg)
```

## Flash initialization (backfit)

### Initial values

```{r flb1}
fl <- fl %>% flash.backfit() %>% flash.nullcheck(remove = TRUE)
plot_loadings(fl$flash.fit$EF[[1]], labs)
```

### Drift results

```{r flb2}
drift.flb <- drift(init_from_flash(fl), miniter = 1000, maxiter = 1000, tol = 0.0005, verbose = FALSE)
ggplot(drift.flb$iterations, aes(x = iter, y = elbo)) + geom_line()
drift.flb[c("elbo", "KL_l", "KL_f")]

plot_loadings(drift.flb$EL, labs, paste("s2:", round(drift.flb$prior_s2, 2)))
plot_cov(drift.flb)
```

<!-- ## Hclust initialization -->

<!-- ### Initial values -->

<!-- ```{r hclust1} -->
<!-- init <- init_using_hclust(simple.admix$Y, k = 4) -->
<!-- plot_loadings(init$EL, labs) -->
<!-- ``` -->

<!-- ### Drift results -->

<!-- ```{r hclust2} -->
<!-- drift.hclust <- drift(init, miniter = 1000, maxiter = 1000, tol = 0.0005, verbose = FALSE) -->
<!-- ggplot(drift.hclust$iterations, aes(x = iter, y = elbo)) + geom_line() -->
<!-- drift.hclust[c("elbo", "KL_l", "KL_f")] -->

<!-- lblr <- paste("s2:", round(drift.hclust$prior_s2, 2)) -->
<!-- names(lblr) <- 1:length(drift.hclust$prior_s2) -->
<!-- lblr <- as_labeller(lblr) -->
<!-- plot_loadings(drift.hclust$EL, labs, lblr) -->
<!-- plot_cov(drift.hclust$EL * rep(sqrt(drift.hclust$prior_s2), each = 80)) -->
<!-- ``` -->

## Initialization from "true" solution

### Initial values

```{r true1}
# I can't give init_from_EL a singular matrix, so I need to fudge the loadings a bit.
EL <- cbind(c(rep(1, 40), rep(0.25, 20), rep(0, 20)),
            c(rep(1, 20), rep(0, 60)),
            c(rep(0, 20), rep(1, 20), rep(0.5, 20), rep(0, 20)),
            c(rep(0, 40), rep(0.5, 20), rep(1, 20)))
init <- init_from_EL(simple.admix$Y, EL)
plot_loadings(init$EL, labs)
```

### Drift results

```{r true2}
drift.true <- drift(init, miniter = 1000, maxiter = 1000, tol = 0.0005, verbose = FALSE)
ggplot(drift.true$iterations, aes(x = iter, y = elbo)) + geom_line()
drift.true[c("elbo", "KL_l", "KL_f")]

plot_loadings(drift.true$EL, labs, paste("s2:", round(drift.true$prior_s2, 2)))
plot_cov(drift.true)
```

## Initialization using three factors

### Initial values

```{r true3}
EL <- cbind(c(rep(1, 20), rep(0, 60)),
            c(rep(0, 20), rep(1, 20), rep(0.5, 20), rep(0, 20)),
            c(rep(0, 40), rep(0.5, 20), rep(1, 20)))
init <- init_from_EL(simple.admix$Y, EL)
plot_loadings(init$EL, labs)
```

### Drift results

```{r true4}
drift.3factor <- drift(init, miniter = 1000, maxiter = 1000, tol = 0.0005, verbose = FALSE)
ggplot(drift.3factor$iterations, aes(x = iter, y = elbo)) + geom_line()
drift.3factor[c("elbo", "KL_l", "KL_f")]

plot_loadings(drift.3factor$EL, labs, paste("s2:", round(drift.3factor$prior_s2, 2)))
plot_cov(drift.3factor)
```

<!-- ## Random initialization -->

<!-- ### Initial values -->

<!-- ```{r rand1} -->
<!-- init <- init_using_hclust(simple.admix$Y, k = 4, as_tree = FALSE) -->
<!-- init$EL <- matrix(rnorm(80 * 4), ncol = 4) -->
<!-- init$EL2 <- init$EL^2 -->
<!-- plot_loadings(init$EL, labs) -->
<!-- ``` -->

<!-- ### Drift results -->

<!-- ```{r rand2} -->
<!-- drift.rand <- drift(init, miniter = 1000, maxiter = 1000, tol = 0.0005, verbose = FALSE) -->
<!-- ggplot(drift.rand$iterations, aes(x = iter, y = elbo)) + geom_line() -->
<!-- drift.rand[c("elbo", "KL_l", "KL_f")] -->

<!-- lblr <- paste("s2:", round(drift.rand$prior_s2, 2)) -->
<!-- names(lblr) <- 1:length(drift.rand$prior_s2) -->
<!-- lblr <- as_labeller(lblr) -->
<!-- plot_loadings(drift.rand$EL, labs, lblr) -->
<!-- plot_cov(drift.rand$EL * rep(sqrt(drift.rand$prior_s2), each = 80)) -->
<!-- ``` -->

<!-- ## One factor per population -->

<!-- ### Initial values -->

<!-- ```{r perpop1} -->
<!-- init <- init_using_hclust(simple.admix$Y, k = 4, as_tree = FALSE) -->
<!-- plot_loadings(init$EL, labs) -->
<!-- ``` -->

<!-- ### Drift results -->

<!-- ```{r perpop2} -->
<!-- drift.perpop <- drift(init, miniter = 1000, maxiter = 1000, tol = 0.0005, verbose = FALSE) -->
<!-- ggplot(drift.perpop$iterations, aes(x = iter, y = elbo)) + geom_line() -->
<!-- drift.perpop[c("elbo", "KL_l", "KL_f")] -->

<!-- lblr <- paste("s2:", round(drift.perpop$prior_s2, 2)) -->
<!-- names(lblr) <- 1:length(drift.perpop$prior_s2) -->
<!-- lblr <- as_labeller(lblr) -->
<!-- plot_loadings(drift.perpop$EL, labs, lblr) -->
<!-- plot_cov(drift.perpop$EL * rep(sqrt(drift.perpop$prior_s2), each = 80)) -->
<!-- ``` -->

## Results summary

```{r summ}
all.drift <- list(drift.flg, drift.flb, drift.true, drift.3factor)

res.df <- data.frame(
  Name = c("flash.greedy", "flash.backfit", "true.4factor", "three.factors"),
  InitialELBO = sapply(all.drift, function(x) x$iterations$elbo[1]),
  FinalELBO = sapply(all.drift, function(x) x$elbo),
  ELBOdiff = sapply(all.drift, function(x) x$elbo - x$iterations$elbo[1]),
  n_iter = sapply(all.drift, function(x) max(x$iterations$iter)),
  KL_l = sapply(all.drift, function(x) sum(x$KL_l)),
  KL_f = sapply(all.drift, function(x) x$KL_f),
  ResidS2 = sapply(all.drift, function(x) x$resid_s2)
)

knitr::kable(res.df, digits = 3)
```

