---
title: "AmericanAdmixture_4B11 [fixed shared factor]"
author: "Joseph Marcus"
date: "2020-05-18"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Here I visualize population structure with simulated data from the [AmericanAdmixture_4B11](https://stdpopsim.readthedocs.io/en/latest/catalog.html#sec_catalog_homsap_models_americanadmixture_4b11) scenario. See [Browning et al. 2018](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1007385) for details.

Below, I show a number of EBMF solutions and in each of them I don't display the first shared factor which is prefixed to the one-vector and scale the loadings by the prior variance. I only describe the loadings that remain after the shared factor.

# Imports

Import the required libraries and scripts:

```{r}
suppressMessages({
  library(lfa)
  library(flashier)
  library(drift.alpha)
  library(ggplot2)
  library(RColorBrewer)
  library(reshape2)
  library(tidyverse)
  library(alstructure)
  source("../code/structure_plot.R")
})
```

## Data

```{r}
data_path <- "../output/simulations/AmericanAdmixture_4B11/rep2.txt"
G <- t(as.matrix(read.table(data_path, sep=" ")))
colnames(G) <- NULL
rownames(G) <- NULL
n <- nrow(G)
daf <- colSums(G) / (2 * n)
colors <- brewer.pal(8, "Set2")

# filter out too rare and too common SNPs
Y <- G[,((daf>=.05) & (daf <=.95))]
p <- ncol(Y)
print(n)
print(p)

# sub-population labels from stdpop
labs <- rep(c("AFR", "EUR", "ASIA", "ADMIX"), each=40)
```

we end up with 160 individuals and ~25000 SNPs. View fitted the sample covariance matrix:

```{r}
plot_cov((1.0 / p) * Y %*% t(Y), as.is=T)
```

# flash [greedy]

Run the `greedy` algorithm:

```{r}
ones <- matrix(1, nrow = n, ncol = 1)
ls.soln <- t(solve(crossprod(ones), crossprod(ones, Y)))
fl <- flash.init(Y) %>%
  flash.init.factors(EF = list(ones, ls.soln), 
                     prior.family=c(prior.bimodal(), prior.normal())) %>%
  flash.fix.loadings(kset = 1, mode = 1L) %>%
  flash.backfit() %>%
  flash.add.greedy(Kmax=8, prior.family=c(prior.bimodal(), prior.normal()))

sd <- unlist(lapply(fl$fitted.g[[2]], '[[', 3))
L <- fl$flash.fit$EF[[1]]
LDsqrt <- L %*% diag(sd)
K <- ncol(LDsqrt)
plot_loadings(LDsqrt[,2:K], labs) + scale_color_brewer(palette="Set2")
```

view structure plot:

```{r}
create_structure_plot(L=LDsqrt[,2:K], labels=labs, colors=colors)
```

view fitted covariance matrix:

```{r}
plot_cov(LDsqrt %*% t(LDsqrt), as.is=T)
```

the `greedy` algorithm picks up a bit of a signal of admixture but misses out on the African contribution.

# flash [backfit]

Run `flash [backfit]` initializing from the greedy solution:

```{r}
flbf <- fl %>% 
  flash.backfit() %>% 
  flash.nullcheck(remove=TRUE)

sd <- unlist(lapply(flbf$fitted.g[[2]], '[[', 3))
L <- flbf$flash.fit$EF[[1]]
LDsqrt <- L %*% diag(sd)
K <- ncol(LDsqrt)
plot_loadings(LDsqrt[,2:K], labs) + scale_color_brewer(palette="Set2")
```

view structure plot:

```{r}
create_structure_plot(L=LDsqrt[,2:K], labels=labs, colors=colors)
```

view fitted covariance matrix:

```{r}
plot_cov(LDsqrt %*% t(LDsqrt), as.is=T)
```

The backfitting algorithm misses out on the signal of admixture and finds only 3 population specific factors after the shared factor.

# drift

Run `drift` initializing from the greedy solution:

```{r}
init <- init_from_data(Y, Kmax=6)
dr <- drift(init, miniter=2, 
            maxiter=1000, 
            tol=0.01, 
            verbose=FALSE)

sd <- sqrt(dr$prior_s2)
L <- dr$EL
LDsqrt <- L %*% diag(sd)
K <- ncol(LDsqrt)
plot_loadings(LDsqrt[,2:K], labs) + scale_color_brewer(palette="Set2")
```

view structure plot:

```{r}
create_structure_plot(L=LDsqrt[,2:K], labels=labs, colors=colors)
```

view fitted covariance matrix:

```{r}
plot_cov(LDsqrt %*% t(LDsqrt), as.is=T)
```

`drift` much better represents the admixture signal but misses the EUR/ASIA two population factor.