---
title: "AmericanAdmixture_4B11"
author: "Joseph Marcus"
date: "2020-05-13"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Here I visualize population structure with simulated data from the [AmericanAdmixture_4B11](https://stdpopsim.readthedocs.io/en/latest/catalog.html#sec_catalog_homsap_models_americanadmixture_4b11) scenario. See [Browning et al. 2018](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1007385) for details.

# Imports

Import the required libraries and scripts:

```{r imports}
suppressMessages({
  library(lfa)
  library(flashier)
  library(drift.alpha)
  library(ggplot2)
  library(reshape2)
  library(tidyverse)
  library(RColorBrewer)
  library(alstructure)
  source("../code/structure_plot.R")
})
```

# Data

```{r data}
data_path <- "../output/simulations/AmericanAdmixture_4B11/rep1.txt"
Y <- t(as.matrix(read.table(data_path, sep=" ")))
n <- nrow(Y)
maf <- colSums(Y) / (2*n)
colors <- brewer.pal(8, "Set2")

# filter out too rare and too common SNPs
Y <- Y[,((maf>=.05) & (maf <=.95))]
p <- ncol(Y)
Z <- scale(Y)
print(n)
print(p)

# sub-population labels from stdpop
labs <- rep(c("AFR", "EUR", "ASIA", "ADMIX"), each=40)
```

we end up with 160 individuals and ~24643 SNPs.

# PCA

Lets run `PCA` on the centered and scaled genotype matrix:

```{r pca}
svd_res <- lfa:::trunc.svd(Z, 5)
L_hat <- svd_res$u
plot_loadings(L_hat, labs) + scale_color_brewer(palette="Set2")
```

Plot the first two factors against each other:

```{r pca-biplot}
qplot(L_hat[,1], L_hat[,2], color=labs) + 
  xlab("PC1") + 
  ylab("PC2") + 
  scale_color_brewer(palette="Set2") + 
  theme_bw()
```

the admixed population is in the center of the PC1 vs PC2 bi-plot.

# ALStructure

Run `ALStructure` with $K=3$:

```{r psd}
admix_res <- alstructure::alstructure(t(Y), d_hat=3)
Qhat <- t(admix_res$Q_hat)
plot_loadings(Qhat, labs) + scale_color_brewer(palette="Set2")
```

view structure plot:

```{r}
create_structure_plot(L=Qhat, labels=labs, colors=colors, ymax=1.01)
```

the three factors seems to represent "ancestral populations" from ASIA, EUR, and AFR and the admixed population draws ancestry from all three of them as expected.

# flash [greedy]

Run the greedy algorithm:

```{r greedy}
fl <- flash(Y, 
            greedy.Kmax=8, 
            prior.family=c(prior.bimodal(), prior.normal()))
plot_loadings(fl$flash.fit$EF[[1]], labs) + scale_color_brewer(palette="Set2")
```

view structure plot:

```{r}
create_structure_plot(L=fl$flash.fit$EF[[1]], labels=labs, colors=colors)
```

there are 4 factors learned by the greedy algorithm: one share factor with a slightly lower loading in the AFR population and 3 sparser factors. The third factor seems to be defined by ASIA individuals and the admixed population has a small weight on this factor.

# flash [backfit]

Run `flash [backfit]` initializing from the greedy solution:

```{r backfit}
flbf <- fl %>% 
  flash.backfit() %>% 
  flash.nullcheck(remove=TRUE)
plot_loadings(flbf$flash.fit$EF[[1]], labs)
```

view structure plot:

```{r}
create_structure_plot(L=flbf$flash.fit$EF[[1]], labels=labs, colors=colors)
```

the results looks qualitatively similar to the greedy algorithm though now the the third factor is completely defined ASIA individuals.

# drift

Run `drift` initializing from the greedy solution:

```{r drift}
init <- init_from_flash(fl)
dr <- drift(init, miniter=2, maxiter=500, tol=0.01, verbose=TRUE)
plot_loadings(dr$EL, labs)
```

view structure plot:

```{r}
create_structure_plot(L=dr$EL, labels=labs, colors=colors)
```

the `drift` algorithm seems to give a pretty different result from `flash`. There is one shared factor but then the admixed population has intermediate values on all the other factors. Here are the admixture proportions from the coalescent simulation:

```
ADMIX percentage 1/6 | Amount African admixture
ADMIX percentage 1/3 | Amount European admixture
ADMIX percentage 1/2 | Amount Asian admixture
```

which actually qualitatively lines up with ADMIX population loadings on factors 2-4! It's kinda interesting that we recapitulate a similar behavior we saw in real data for "known" admixed populations when we applied `flash` and `drift` to Human Origins data.