---
title: "Fitting a real-data covariance matrix: details"
author: "Jason Willwerscheid"
date: "9/11/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 6.5)
```

```{r load.pkgs}
suppressMessages({
  library(flashier)
  library(drift.alpha)
  library(tidyverse)
})
```

I add a few fits and more detailed plots to the [previous analysis](covmat_1kg_shared.html).

```{r dat}
covmat <- readRDS("../data/datasets/1kg_phase3_derived/1kg_phase3_derived_covmat.rds")
meta <- readRDS("../data/datasets/1kg_phase3_derived/1kg_phase3_derived_meta.rds")

plot_fl <- function(LL) {
  df <- data.frame(LL)
  colnames(df) <- paste0("Factor ", formatC(1:ncol(LL), width = 2, flag = "0"))
  df$subpop <- meta$pop
  df$superpop <- meta$super_pop
  df <- df %>% arrange(superpop, subpop)
  df$idx <- 1:nrow(df)
  gath_df <- df %>% 
    gather(K, value, -subpop, -idx, -superpop) %>%
    mutate(K = factor(K))
  med_gath_df <- gath_df %>% 
    group_by(subpop, K) %>% 
    summarise(value=median(value), idx=median(idx))
  
  p <- ggplot(gath_df, aes(x=idx, y=value, color=superpop)) + 
    geom_point() +
    facet_wrap(~K) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    labs(color="superpop")
  return(p)
}

plot_subpops <- function(LL) {
  df <- data.frame(LL)
  colnames(df) <- paste0("Factor ", formatC(1:ncol(LL), width = 2, flag = "0"))
  df$subpop <- meta$pop
  df$superpop <- meta$super_pop
  df <- df %>% arrange(superpop, subpop)
  df$idx <- 1:nrow(df)
  gath_df <- df %>% 
    gather(K, value, -subpop, -idx, -superpop) %>%
    mutate(K = factor(K))
  med_gath_df <- gath_df %>% 
    group_by(subpop, K) %>% 
    summarise(value=median(value), idx=median(idx))
  
  all_plots <- lapply(levels(df$superpop), function(pop) {
    p <- ggplot(filter(gath_df, superpop == pop), aes(x=idx, y=value, color=subpop)) + 
      geom_point() +
      facet_wrap(~K) + 
      geom_hline(yintercept = 0, linetype = "dashed") +
      theme(axis.title.x=element_blank(),
            axis.text.x=element_blank(),
            axis.ticks.x=element_blank()) +
      labs(color="subpop") +
      ggtitle(paste("superpop:", pop))
    return(p)  
  })
}
```

## Greedy fit (point-Laplace)

```{r pl_g}
ones <- matrix(1, nrow = nrow(covmat), ncol = 1)
ls.soln <- t(solve(crossprod(ones), crossprod(ones, covmat)))

diag(covmat) <- NA

fl_g <- flash.init(covmat) %>%
  flash.set.verbose(0) %>%
  flash.init.factors(EF = list(ones, ls.soln)) %>%
  flash.fix.loadings(kset = 1, mode = 1) %>%
  flash.backfit() %>%
  flash.add.greedy(Kmax = 11, 
                   prior.family = prior.point.laplace())

plot(plot_fl(fl_g$loadings.pm[[1]] %*% diag(sqrt(fl_g$loadings.scale))))
```

## Backfit (point-Laplace)

```{r pl_bf}
fl_bf <- fl_g %>% flash.backfit(maxiter = 1000)

plot(plot_fl(fl_bf$loadings.pm[[1]] %*% diag(sqrt(fl_bf$loadings.scale))))
```

## Drift (point-Laplace, $p = 100000$)

The true value of $p$ is around 100k:

```{r dr}
covmat <- readRDS("../data/datasets/1kg_phase3_derived/1kg_phase3_derived_covmat.rds")
LL <- fl_bf$loadings.pm[[1]] %*% diag(sqrt(fl_bf$loadings.scale))

dr <- init_from_covmat(covmat, LL, p = 100000, fix.EL = 1,
                       prior.family = prior.point.laplace())
dr <- drift(dr, maxiter = 500, tol = 1e-4, verbose = FALSE)

plot(plot_fl(dr$EL %*% diag(sqrt(dr$prior_s2))))
```

## Drift (point-Laplace, $p = 10000$)

I can set $p$ smaller to get a sparser solution:

```{r dr_10000}
dr10000 <- init_from_covmat(covmat, LL, p = 10000, fix.EL = 1,
                       prior.family = prior.point.laplace())
dr10000 <- drift(dr10000, maxiter = 500, tol = 1e-4, verbose = FALSE)

plot(plot_fl(dr10000$EL %*% diag(sqrt(dr10000$prior_s2))))
```

## Subpopulations (point-Laplace drift)

```{r subpops}
all_p <- plot_subpops(dr10000$EL %*% diag(sqrt(dr10000$prior_s2)))
for (p in all_p) {plot(p)}
```

## Drift (bimodal)

I initialize using the sparser factors from `drift` with $p = 10000$:

```{r bimodal}
LL <- dr10000$EL[, c(1, 2, 2, 3, 3, 4, 5, 6, 6, 8, 8, 9, 10, 11, 12)]
LL <- t(t(LL) * c(1, 1, -1, 1, -1, -1, -1, 1, -1, 1, -1, -1, -1, 1, -1))
LL <- pmax(LL, 0)

LL_scale <- apply(LL, 2, max)
LL <- t(t(LL) / LL_scale)

dr_bimodal <- init_from_covmat(covmat, LL, p = 10000, prior.s2 = LL_scale^2, fix.EL = 1,
                               prior.family = prior.bimodal())
dr_bimodal <- drift(dr_bimodal, maxiter = 500, tol = 1e-4, verbose = FALSE)

plot(plot_fl(dr_bimodal$EL %*% diag(sqrt(dr_bimodal$prior_s2))))
```

## Subpopulations (bimodal drift)

```{r subpops_bimodal}
all_p <- plot_subpops(dr_bimodal$EL %*% diag(sqrt(dr_bimodal$prior_s2)))
for (p in all_p) {plot(p)}
```
