---
title: "Fitting a real-data covariance matrix: interpretation"
author: "Jason Willwerscheid"
date: "9/11/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 8, fig.height = 6.5)
```

```{r load.pkgs}
suppressMessages({
  library(flashier)
  library(drift.alpha)
  library(tidyverse)
})
```

I interpret results from the [previous analysis](covmat_1kg_shared2.html). I do a bimodal drift fit and then sort factors by the L2 norm of their loadings.

## Superpopulations

```{r dat}
covmat <- readRDS("../data/datasets/1kg_phase3_derived/1kg_phase3_derived_covmat.rds")
meta <- readRDS("../data/datasets/1kg_phase3_derived/1kg_phase3_derived_meta.rds")

plot_fl <- function(LL) {
  df <- data.frame(LL)
  colnames(df) <- paste0("Factor ", formatC(1:ncol(LL), width = 2, flag = "0"))
  df$subpop <- meta$pop
  df$superpop <- meta$super_pop
  df <- df %>% arrange(superpop, subpop)
  df$idx <- 1:nrow(df)
  gath_df <- df %>% 
    gather(K, value, -subpop, -idx, -superpop) %>%
    mutate(K = factor(K))
  med_gath_df <- gath_df %>% 
    group_by(subpop, K) %>% 
    summarise(value=median(value), idx=median(idx))
  
  p <- ggplot(gath_df, aes(x=idx, y=value, color=superpop)) + 
    geom_point() +
    facet_wrap(~K) + 
    geom_hline(yintercept = 0, linetype = "dashed") +
    theme(axis.title.x=element_blank(),
          axis.text.x=element_blank(),
          axis.ticks.x=element_blank()) +
    labs(color="superpop")
  return(p)
}

plot_subpops <- function(LL) {
  df <- data.frame(LL)
  colnames(df) <- paste0("Factor ", formatC(1:ncol(LL), width = 2, flag = "0"))
  df$subpop <- meta$pop
  df$superpop <- meta$super_pop
  df <- df %>% arrange(superpop, subpop)
  df$idx <- 1:nrow(df)
  gath_df <- df %>% 
    gather(K, value, -subpop, -idx, -superpop) %>%
    mutate(K = factor(K))
  med_gath_df <- gath_df %>% 
    group_by(subpop, K) %>% 
    summarise(value=median(value), idx=median(idx))
  
  all_plots <- lapply(levels(df$superpop), function(pop) {
    p <- ggplot(filter(gath_df, superpop == pop), aes(x=idx, y=value, color=subpop)) + 
      geom_point() +
      facet_wrap(~K) + 
      geom_hline(yintercept = 0, linetype = "dashed") +
      theme(axis.title.x=element_blank(),
            axis.text.x=element_blank(),
            axis.ticks.x=element_blank()) +
      labs(color="subpop") +
      ggtitle(paste("superpop:", pop))
    return(p)  
  })
}
```

```{r pl_g}
t_greedy <- system.time({
  ones <- matrix(1, nrow = nrow(covmat), ncol = 1)
  ls.soln <- t(solve(crossprod(ones), crossprod(ones, covmat)))
  
  covmat_diagNA <- covmat
  diag(covmat_diagNA) <- NA
  
  fl_g <- flash.init(covmat_diagNA) %>%
    flash.set.verbose(0) %>%
    flash.init.factors(EF = list(ones, ls.soln)) %>%
    flash.fix.loadings(kset = 1, mode = 1) %>%
    flash.backfit() %>%
    flash.add.greedy(Kmax = 11, 
                     prior.family = prior.point.laplace())
})

cat("Time to fit greedy factors:", round(t_greedy[3], 1), "seconds")
```

```{r pl_bf}
t_backfit <- system.time({
  fl_bf <- fl_g %>% flash.backfit(maxiter = 1000)
})

cat("Time to backfit:", round(t_backfit[3] / 60, 1), "minutes")
```

```{r dr_10000}
LL <- fl_bf$loadings.pm[[1]] %*% diag(sqrt(fl_bf$loadings.scale))
LL <- LL[, c(1, rep(2:fl_bf$n.factors, each = 2))]
LL <- t(t(LL) * c(1, rep(c(1, -1), fl_bf$n.factors - 1)))
LL <- pmax(LL, 0)

LL_scale <- apply(LL, 2, max)
LL <- t(t(LL) / LL_scale)

t_drift <- system.time({
  dr <- init_from_covmat(covmat, LL, p = 10000, prior.s2 = LL_scale^2, fix.EL = 1,
                         prior.family = prior.bimodal())
  dr <- drift(dr, maxiter = 500, tol = 1e-4, verbose = FALSE)
})

cat("Time to drift:", round(t_drift[3] / 60, 1), "minutes")

LL <- dr$EL %*% diag(sqrt(dr$prior_s2))
LL_norms <- apply(LL, 2, function(x) sum(x^2))
LL <- LL[, order(LL_norms, decreasing = TRUE)]
LL_norms <- apply(LL, 2, function(x) sum(x^2))
zero_cols <- which(LL_norms < 1e-6)
LL <- LL[, -zero_cols]

plot(plot_fl(LL))
```

## Subpopulations

```{r subpops}
all_p <- plot_subpops(LL)
for (p in all_p) {plot(p)}
```

## Factor descriptions

* Factor 2: The primary African factor, with varying degrees of admixture for African Americans (ASW),  African Caribbeans (ACB), and Amerindians. Intriguingly, there is a small degree of admixture for several Iberians (IBR) and Gujarati Indians from Houston (GIH).

* Factor 3: The primary East Asian factor. There is a very small amount of admixture for Peruvians (PEL) as well as for a small number of African Americans, Mexicans (MXL), and Finns (FIN). A small degree of admixture is common among Bengali (BEB) and Telugu (ITU) Indians and Sri Lankans (STU), and present but less common among Gujarati Indians.

* Factor 4: The primary European factor, with strong contributions from South Asian populations (especially Punjabi and Gujarati) and varying degrees of admixture for African Americans, African Caribbeans, and Amerindians. There is a single Vietnamese (KHV) individual with a nonzero loading: this can probably be ignored.

* Factor 5: The primary South Asian factor. There is a single African Caribbean individual with a nonzero loading.

* Factor 6: The Amerindian factor, with loadings largest among Peruvians and smallest among Puerto Ricans (PUR). Several African Caribbeans are loaded on this factor, as well as, interestingly, a few Gujarati Indians from Houston.

* Factor 7: The out-of-Africa factor. Oddly, it is shared by Mexicans, Peruvians, and most Columbians (CLM), but not by most Puerto Ricans.

* Factor 8: Substructure among East Asians: large loadings for the more Southeastern Vietnamese and Chinese Dai populations, modest loadings for some Chinese Han, and mostly zero loadings for Japanese.

* Factor 9: A West African factor: large loadings for Gambians (GWD) and modest loadings for the Mende from Sierra Leone (MSL).

* Factor 10: Shared by all Amerindian and European populations except for Toscani. I don't know where this comes from.

* Factor 11: A Mediterranean factor, possibly? Shared by Southern European (Iberian and Toscani) and Puerto Rican populations and a large number of Mexicans and Colombians.

* Factor 12: A Finnish factor, also present among one or two Utah residents with European ancestry (CEU).

* Factor 13: Kenyan.

* Factor 14: Japanese.

* Factor 15: Common to some, but not all Gujarati Indians from Houston.

* Factor 16: Shared among Japanese, most Chinese Han from Beijing, and some Southern Chinese Han.

* Factor 17: Common to some, but not all Punjabi.

* Factor 18: Shared among the two Nigerian populations (Esan (ESN) and Yoruba (YRI)).

* Factor 19: Shared among most Toscani and Iberians and some British and Utah residents. One Chinese Dai individual has a nonzero loading.
