---
title: "Human Origins Array Global Data"
author: "jhmarcus"
date: "2019-03-04"
output: workflowr::wflow_html
---

# Imports

Lets import some needed packages:

```{r imports, warning=FALSE, message=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(lfa)
```

# Read Genotypes

Here I read the full genotype matrix of the Human Origins dataset:

```{r genotypes, warning=FALSE, message=FALSE}
Y = t(lfa:::read.bed("../data/raw/NearEastPublic/HumanOriginsPublic2068"))

# number of individuals
n = nrow(Y)

# number of SNPs
p = ncol(Y)
```

# Missingness per SNP

```{r miss-per-snp}
n_miss_snp = colSums(is.na(Y))
p_snpmss = qplot(n_miss_snp) + geom_histogram() + theme_bw()
p_snpmss
```

There are very few SNPs with high levels of missing data so we can use a very stringent missingness threshold without losing much information.

```{r misssnp-thresh}
sum(n_miss_snp==1)
sum(n_miss_snp==2)
sum(n_miss_snp==3)
sum(n_miss_snp %in% 1:10)
snp_idx = which(n_miss_snp <= 10)
10 / n
```

It seems like .995% is reasonable cutoff for missingness.

# Missingness per Individual

```{r miss-per-ind}
n_miss_ind = rowSums(is.na(Y))
p_indmss = qplot(n_miss_ind) + geom_histogram() + theme_bw()
p_indmss
```

```{r missind-thresh}
sum(n_miss_ind > 20000)
20000 / p
```

It seems like a few individuals are missing about 3% of their SNPs which is a bit worrisome maybe they should be remove from the analysis? For now I will in include them and see if they pop up as any outliers in the PCs.



