---
title: "Empircal Bayes Normal Means with Bernoulli Priors"
author: "jhmarcus"
date: "2019-05-23"
output: workflowr::wflow_html
---

Here I explore solving the Empirical Bayes Normal Means problem with a Bernoulli prior. This could be useful for fitting highly constrained non-negative FLASH.

$$
\beta_j \sim Bern(\pi) \\
y_j | \beta_j \sim N(\beta_j, s^2_j)
$$
The marginal likelihood is given by:

$$
\begin{aligned}
p(y_j |\pi) &= p(\beta_j=0)p(y_j | \beta_j = 0) + p(\beta_j=1) p(y_j|\beta_j =1) \\
&= (1-\pi)p(y_j|\beta_j=0) + \pi p(y_j|\beta_j=1)
\end{aligned}
$$

We can easily maximize the marginal likelihood to estimate $\hat{\pi}$ using a root finding algorithm, such as Brent's method. To compute the posterior mean we can simply use Bayes rule and plug in the estimated prior:

$$
\begin{aligned}
E(\beta_j|y_j) &= p(\beta_j=1|y_j) \\ 
&=\frac{p(y_j|\beta_j=1)p(\beta_j=1)}{p(y_j|\beta_j=0)p(\beta_j=0) + p(y_j | \beta_j=1)p(\beta_j=1)} \\
&= \frac{p(y_j|\beta_j=1)\hat{\pi}}{p(y_j|\beta_j=0)(1-\hat{\pi}) + p(y_j | \beta_j=1)\hat{\pi}}
\end{aligned}
$$

# Imports

```{r warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
```

# Functions

Here are some helper function for simulation, fitting, and plotting.

```{r}
sim = function(n0, n1, sigma_e){
  n = n0+n1
  beta = c(rep(0, n0), rep(1, n1))
  s = abs(rnorm(n, 0, sigma_e))
  betahat = rnorm(n, beta, s)
  return(list(betahat=betahat, s=s, beta=beta, n=n))
}

nll = function(pi, data){
  betahat = data[[1]]
  s = data[[2]]
  liks = (1.0-pi) * dnorm(betahat, mean=0, sd=s) + pi * dnorm(betahat, mean=1, sd=s) 
  loglik = sum(log(liks))
  return(-loglik)
}

est_pi = function(betahat, s){
  fit = optim(par=c(.5), 
              fn=nll, 
              data=list(betahat, s), 
              method="Brent", 
              lower=0, 
              upper=1)

  pihat = fit$par
  return(pihat)
}

est_post = function(betahat, s, pihat){
  c0 = dnorm(betahat, mean = 0, sd=s) * (1 - pihat)
  c1 = dnorm(betahat, mean = 1, sd=s) * pihat
  pm = c1 / (c0 + c1)
  return(pm)
}

plot_sim = function(df, title){
  
  gath_df = df %>% gather(variable, value, -idx, -s)
  p0 = ggplot(gath_df, aes(x=idx, y=value, 
                        color=factor(variable, levels=c("beta", "betahat", "betapm")))) + 
      geom_point() + 
      theme_bw() +
      labs(color="") +
      xlab("Variable") + 
      ylab("Value") +
      theme(legend.position="bottom")
  
  min_betahat = min(df$betahat)
  max_betahat = max(df$betahat)
  p1 = ggplot(df, aes(betahat, betapm, color=s)) + 
       geom_point() + viridis::scale_color_viridis() + 
       theme_bw() + 
       theme(legend.position="bottom") +
       xlim(c(min_betahat, max_betahat)) +
       ylim(c(min_betahat, max_betahat)) + 
       geom_abline() 

  p = cowplot::plot_grid(p0, p1, nrow=1) 
  title = cowplot::ggdraw() + cowplot::draw_label(title)
  print(cowplot::plot_grid(title, p, ncol=1, rel_heights=c(0.1, 1)))
  
}
```

```{r, eval=FALSE}
n0=10
n1=90
sigma_e=.25
sim_res = sim(n0, n1, sigma_e)
betahat = sim_res$betahat
s = sim_res$s
beta = sim_res$beta
pihat = est_pi(betahat, s)
betapm = est_post(betahat, s, pihat)
df = data.frame(betahat=betahat, beta=beta, betapm=betapm, s=s, idx=1:length(betahat))
title = paste0("n0=",n0, ",n1=", n1, ",sigma_e=", sigma_e, ",pihat=", round(pihat, digits=3))
plot_sim(df, title)
```

I simulated a bunch of normal means scenarios where the true $\beta$s are set to 0 or 1. In each simulation I specify the number of zeros `n0` the number of ones `n1` and standard deviation used to simulate std. errors.

```{r}
n0 = c(rep(40, 3), rep(25, 3), rep(10, 3), rep(0, 3))
n1 = c(rep(40, 3), rep(55, 3), rep(70, 3), rep(80, 3))
sigma_e = rep(c(.05, .1, .25), 4)

for(i in 1:length(n0)){
  sim_res = sim(n0[i], n1[i], sigma_e[i])
  betahat = sim_res$betahat
  s = sim_res$s
  beta = sim_res$beta
  pihat = est_pi(betahat, s)
  betapm = est_post(betahat, s, pihat)
  df = data.frame(betahat=betahat, beta=beta, betapm=betapm, s=s, idx=1:length(betahat))
  title = paste0("n0=",n0[i], ",n1=", n1[i], ",sigma_e=", sigma_e[i], ",pihat=", round(pihat, digits=3))
  plot_sim(df, title)
}
```

