---
title: "Extrapolation Test"
author: "Jason Willwerscheid"
date: "6/24/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 5, fig.height = 4)
```

```{r load.pkgs}
suppressMessages({
  library(flashier)
  library(drift.alpha)
  library(tidyverse)
})
```

I test the "extrapolation" method that I [developed](https://willwerscheid.github.io/FLASHvestigations/extrapolate.html) for `flashier` and have since implemented in `drift`. I use the same four-population dataset that I used in a  [previous analysis](tree_adjust.html).

```{r sim}
set.seed(666)

n_per_pop <- 60
p <- 10000

a <- rnorm(p)
b <- rnorm(p)
c <- rnorm(p)
d <- rnorm(p, sd = 0.5)
e <- rnorm(p, sd = 0.5)
f <- rnorm(p, sd = 0.5)
g <- rnorm(p, sd = 0.5)

popA <- c(rep(1, n_per_pop), rep(0, 3 * n_per_pop))
popB <- c(rep(0, n_per_pop), rep(1, n_per_pop), rep(0, 2 * n_per_pop))
popC <- c(rep(0, 2 * n_per_pop), rep(1, n_per_pop), rep(0, n_per_pop))
popD <- c(rep(0, 3 * n_per_pop), rep(1, n_per_pop))

Y <- cbind(popA, popB, popC, popD) %*% rbind(a + b + d, a + b + e, a + c + f, a + c + g)
Y <- Y + rnorm(4 * n_per_pop * p, sd = 0.1)

plot_dr <- function(dr) {
  sd <- sqrt(dr$prior_s2)
  L <- dr$EL
  LDsqrt <- L %*% diag(sd)
  K <- ncol(LDsqrt)
  plot_loadings(LDsqrt[,1:K], rep(c("A", "B", "C", "D"), each = n_per_pop)) +
    scale_color_brewer(palette="Set2")
}
```

The approach works well. The fit that uses extrapolation to accelerate gets to a much better solution fairly quickly:

```{r fits}
# For reasons that I don't yet understand, extrapolation works especially well for
#   relatively small values of beta.max.
options(extrapolate.control = list(beta.max = 1.1))

greedy <- init_from_data(Y)
dr_slow <- drift(greedy, verbose = FALSE, extrapolate = FALSE, maxiter = 2000, tol = 1e-4)
dr_xtrap <- drift(greedy, verbose = FALSE, extrapolate = TRUE, maxiter = 2000, tol = 1e-4)

df <- dr_slow$iterations %>%
  mutate(extrapolate = "FALSE") %>%
  bind_rows(dr_xtrap$iterations %>% mutate(extrapolate = "TRUE"))
ggplot(df, aes(x = iter, y = elbo, col = extrapolate)) + geom_line()
```

The solutions are qualitatively different. Without extrapolation, the solution is highly suboptimal (as I've shown elsewhere):

```{r figs1}
plot_dr(dr_slow)
```

The extrapolated solution is much better. In particular, the tree-like structure of the data is readily apparent:

```{r figs2}
plot_dr(dr_xtrap)
```
