---
title: "Extrapolation Test (Admixture)"
author: "Jason Willwerscheid"
date: "6/25/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 5, fig.height = 4)
```

```{r load.pkgs}
suppressMessages({
  library(flashier)
  library(drift.alpha)
  library(tidyverse)
})
```

I test extrapolation on a four-population tree with an additional admixed population. The latter has admixture proportions of $1/2$ Population B, $1/3$ Population C, and $1/6$ Population D.

```{r sim}
set.seed(666)

n_per_pop <- 60
p <- 10000

a <- rnorm(p)
b <- rnorm(p)
c <- rnorm(p)
d <- rnorm(p, sd = 0.5)
e <- rnorm(p, sd = 0.5)
f <- rnorm(p, sd = 0.5)
g <- rnorm(p, sd = 0.5)

popA <- c(rep(1, n_per_pop), rep(0, 4 * n_per_pop))
popB <- c(rep(0, n_per_pop), rep(1, n_per_pop), rep(0, 3 * n_per_pop))
popC <- c(rep(0, 2 * n_per_pop), rep(1, n_per_pop), rep(0, 2 * n_per_pop))
popD <- c(rep(0, 3 * n_per_pop), rep(1, n_per_pop), rep(0, n_per_pop))
popE <- c(rep(0, 4 * n_per_pop), rep(1, n_per_pop))

E.factor <- (a + b + e) / 2 + (a + c + f) / 3 + (a + c + g) / 6

Y <- cbind(popA, popB, popC, popD, popE) %*% 
  rbind(a + b + d, a + b + e, a + c + f, a + c + g, E.factor)
Y <- Y + rnorm(5 * n_per_pop * p, sd = 0.1)

plot_dr <- function(dr) {
  sd <- sqrt(dr$prior_s2)
  L <- dr$EL
  LDsqrt <- L %*% diag(sd)
  K <- ncol(LDsqrt)
  plot_loadings(LDsqrt[,1:K], rep(c("A", "B", "C", "D", "E"), each = n_per_pop)) +
    scale_color_brewer(palette="Set2")
}
```

Extrapolation again finds a much better solution within 2000 iterations:

```{r fits}
greedy <- init_from_data(Y)

dr_slow <- drift(greedy, verbose = FALSE, extrapolate = FALSE, maxiter = 2000, tol = 1e-4)

options(extrapolate.control = list(beta.max = 1))
dr_xtrap <- drift(greedy, verbose = FALSE, extrapolate = TRUE, maxiter = 2000, tol = 1e-4)

df <- dr_slow$iterations %>%
  mutate(extrapolate = "FALSE") %>%
  bind_rows(dr_xtrap$iterations %>% mutate(extrapolate = "TRUE"))
ggplot(df, aes(x = iter, y = elbo, col = extrapolate)) + geom_line()
```

Without extrapolation, the solution is very difficult to interpret:

```{r figs1}
plot_dr(dr_slow)
```

The extrapolated solution isn't perfect, but it's pretty good. It captures the underlying tree fairly well, and it correctly represents population E as an admixture (even though it's missing the contribution from Population C).

```{r figs2}
plot_dr(dr_xtrap)
```
