---
title: "Alternative view of Human Origins factors"
author: "Peter Carbonetto"
output: workflowr::wflow_html
---

Following from [the initial analysis](hoa_global.html), this analysis
presents an alternative view of the factors.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(fig.align = "center")
```

## Analysis settings

This is the file with a large data frame containing the factor
loadings and other sample information.

```{r set-input-file-names}
loadings.file <- file.path("..","sandbox","loadings-forpeter-03-12-2019.rds")
```

## Set up environment

Load several R packages and function definitions used in the code
chunks below.

```{r load-pkgs, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggstance)
library(cowplot)
source(file.path("..","code","hoa_global_alt_functions.R"))
```

## Load results

Load the data frame containing the factor loadings and population labels.

```{r load-results}
hoa <- load.results(loadings.file)
```

This data frame should contain information on 2,018 genotype samples:

```{r inspect-results-data-frame}
nrow(hoa)
```

## Factors 2--21

The following plots are intended to help interpret the factors by
relating them to the provided population labels.

### Factor 2

This plot shows the median loading by assigned population label, with
error bars capturing the 5th and 95th percentiles. Colours represent
broad geographic groups. Populations in which the largest loading is
less than 0.01 are not shown.

The second factor appears to capture east Asian, Oceanian and American
populations, among others.

```{r plot-factor-2, fig.width=4.5, fig.height=6}
with(hoa,plot.response.by.label(factor2,Simple.Population.ID,Region))
```

### Factor 3

Factor 3 appears to capture mainly sub-Saharan African populations.

```{r plot-factor-3, fig.width=4.5, fig.height=3.5}
with(hoa,plot.response.by.label(factor3,Simple.Population.ID,Region))
```

### Factor 4

Factor 4 seems to capture mainly European and Middle Eastern ancestry.

```{r plot-factor-4, fig.width=4.5, fig.height=8}
with(hoa,plot.response.by.label(factor4,Simple.Population.ID,Region))
```

### Factor 5

Factor 5 captures Papuan and Australian populations.

```{r plot-factor-5, fig.width=4.5, fig.height=4}
with(hoa,plot.response.by.label(factor5,Simple.Population.ID,Region))
```

### Factor 6

Factor 6 is largely capturing South American populations.

```{r plot-factor-6, fig.width=4.5, fig.height=3}
with(hoa,plot.response.by.label(factor6,Simple.Population.ID,Region))
```

### Factor 7

Factor 7 seems to reflect East Asian origins.

```{r plot-factor-7, fig.width=4.5, fig.height=6}
with(hoa,plot.response.by.label(factor7,Simple.Population.ID,Region))
```

### Factor 8

Factor 8 is some combination of populations originating in Siberia and
Russia.

```{r plot-factor-8, fig.width=4.5, fig.height=6}
with(hoa,plot.response.by.label(factor8,Simple.Population.ID,Region))
```

### Factor 9

Factor 9 corresponds largely to populations from India, as well as
Middle Eastern and Central Eurasian groups.

```{r plot-factor-9, fig.width=4.5, fig.height=5.5}
with(hoa,plot.response.by.label(factor9,Simple.Population.ID,Region))
```

### Factor 10

Factor 10 picks up a small number groups from sub-Saharan Africa,
including the Khomani and Mbuti.

```{r plot-factor-10, fig.width=4.5, fig.height=2}
with(hoa,plot.response.by.label(factor10,Simple.Population.ID,Region))
```

### Factor 11

Factor 11 is capturing some subset of Saharan and Middle Eastern
populations.

```{r plot-factor-11, fig.width=4.5, fig.height=6}
with(hoa,plot.response.by.label(factor11,Simple.Population.ID,Region))
```

### Factor 12

Factor 12 seems to distinguish European populations.

```{r plot-factor-12, fig.width=4.5, fig.height=8}
with(hoa,plot.response.by.label(factor12,Simple.Population.ID,Region))
```

### Factor 13

Interpreting Factor 13 requires some better understanding of the
Nganasan, Dolgan and other groups.

```{r plot-factor-13, fig.width=4.5, fig.height=6}
with(hoa,plot.response.by.label(factor13,Simple.Population.ID,Region))
```

### Factor 14

Factor 14 is difficult to interpret.

```{r plot-factor-14, fig.width=4.5, fig.height=6}
with(hoa,plot.response.by.label(factor14,Simple.Population.ID,Region))
```

### Factor 15

Factor 15 is also difficult to interpret.

```{r plot-factor-15, fig.width=4.5, fig.height=6}
with(hoa,plot.response.by.label(factor15,Simple.Population.ID,Region))
```

### Factor 16

Factor 16 captures individuals of Surui origin.

```{r plot-factor-16, fig.width=4.5, fig.height=2}
with(hoa,plot.response.by.label(factor16,Simple.Population.ID,Region))
```

### Factor 17

Factor 17 captures the Biaka ancestral population.

```{r plot-factor-17, fig.width=4.5, fig.height=2.5}
with(hoa,plot.response.by.label(factor17,Simple.Population.ID,Region))
```

### Factor 18

Factor 18 picks out people with Karitianan origins.

```{r plot-factor-18, fig.width=4.5, fig.height=2.5}
with(hoa,plot.response.by.label(factor18,Simple.Population.ID,Region))
```

### Factor 19

Factor 19 captures the Pima ancestral population.

```{r plot-factor-19, fig.width=4.5, fig.height=2}
with(hoa,plot.response.by.label(factor19,Simple.Population.ID,Region))
```

### Factor 20

Factor 20 captures a small subset of central Asian populations,
including Itelmen and Koryak.

```{r plot-factor-20, fig.width=4.5, fig.height=7}
with(hoa,plot.response.by.label(factor20,Simple.Population.ID,Region))
```

### Factor 21

Factor 21 picks out the Mbuti.

```{r plot-factor-21, fig.width=4.5, fig.height=2}
with(hoa,plot.response.by.label(factor21,Simple.Population.ID,Region))
```
