---
title: "Alternative view of Human Origins factors"
author: "Peter Carbonetto"
output: workflowr::wflow_html
---

Following from [the initial analysis](hoa_global.html), this analysis
presents an alternative view of the factors.

## Analysis settings

This is the file with a large data frame containing the factor
loadings and other sample information.

```{r set-input-file-names}
loadings.file <- file.path("..","sandbox","loadings-forpeter-03-12-2019.rds")
```

## Set up environment

Load several R packages and function definitions used in the code
chunks below.

```{r load-pkgs, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggstance)
library(cowplot)
source(file.path("..","code","hoa_global_alt_functions.R"))
```

## Load results

Load the data frame containing the factor loadings and population labels.

```{r load-results}
hoa <- load.results(loadings.file)
```

This data frame should contain information on 2,018 genotype samples:

```{r inspect-results-data-frame}
nrow(hoa)
```

## Factors 2--21

*Add text here.*

### Factor 2

### Factor 3

### Factor 4

### Factor 5

### Factor 6

### Factor 7

### Factor 8

### Factor 9

### Factor 10
