---
title: "migration_sim"
author: "jhmarcus"
date: "2019-04-04"
output: workflowr::wflow_html
---

Here I analze simulated genotype data from a tree. See [Pickrell et al. 2012 Figure 2](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1002967) for details on the the demographic model used.

# Imports

Lets import some needed packages:

```{r imports, warning=FALSE, message=FALSE}
library(ggplot2)
library(ggridges)
library(tidyr)
library(dplyr)
library(RColorBrewer)
library(biomaRt)
library(knitr)
source("../code/viz.R")
source("../code/flash_prep.R")
```

# FLASH-Greedy

Lets first read the greedy `flashier` fit

```{r flash-greedy}
# read the flash fit output by snakemake
flash_fit = readRDS("../output/flash_backfit/migration_sim/migration_sim_maf.rds")
K = flash_fit$n.factors
n = nrow(flash_fit$loadings$normalized.loadings[[1]])
p = nrow(flash_fit$loadings$normalized.loadings[[2]])

snp_df = prepare_snp_data(flash_fit, "../data/datasets/migration_sim/migration_sim_maf.bim")
pves = flash_fit$pve[2:K]

print(K)
print(n)
print(p)
```

# PVEs
 
Here is a plot of the proportion of variance (PVE) explained by each drift event:

```{r flash-greedy-ld-viz-pve, warning=FALSE, message=FALSE}
plot_pve(flash_fit)
```

# Distribution of Factors

```{r flash-greedy-ld-viz-pve, warning=FALSE, message=FALSE}
plot_factors(snp_df, rel_size=.7) 
```

# Fitted Mean and Variance

I setup the `flashier` run so it estimates a SNP specific precision term. Here is a histogram of fitted variances:

```{r flash-greedy-viz-var, warning=FALSE, message=FALSE}
plot_variance(snp_df)
```

Lets now look the the fitted means:

```{r flash-greedy-viz-mean, warning=FALSE, message=FALSE}
p_m = plot_mean(snp_df)
p_v = plot_variance(snp_df)
p_mv = plot_mean_variance(snp_df)

cowplot::plot_grid(p_m, p_v, p_mv, nrow = 2)

#p = cowplot::plot_grid(p_africa, p_west_eurasia, p_central_asia_siberia,
#                       p_america, p_east_asia, p_south_asia, p_oceania, 
#                       rel_heights = c(1.2, 1.3, 1, 1, 1, 1, 1.1),
#                       nrow = 7, align = "v") 
```

It looks like many of the SNPs are on the rarer side. We should see a quadratic relationship with the estimated variance:

```{r flash-greedy-viz-mv, warning=FALSE, message=FALSE}
plot_mean_variance(snp_df)
```

```{r flash-viz-loadings, warning=FALSE, message=FALSE}
# setup loadings data.frame
l_df = as.data.frame(flash_fit$loadings$normalized.loadings[[1]])
l_df$iid = 1:200
l_df$clst = c(rep("pop1", 10), 
              rep("pop2", 10), 
              rep("pop3", 10), 
              rep("pop4", 10), 
              rep("pop5", 10), 
              rep("pop6", 10), 
              rep("pop7", 10), 
              rep("pop8", 10), 
              rep("pop9", 10), 
              rep("pop10", 10), 
              rep("pop11", 10), 
              rep("pop12", 10), 
              rep("pop13", 10), 
              rep("pop14", 10), 
              rep("pop15", 10), 
              rep("pop16", 10), 
              rep("pop17", 10), 
              rep("pop18", 10), 
              rep("pop19", 10), 
              rep("pop20", 10))
colnames(l_df)[1:32] = 1:32
pops = paste0("pop", 1:20)

# gather the data.frame for plotting
gath_df = l_df %>%
          select_if(~sum(!is.na(.)) > 0) %>%
          gather(K, value, -iid, -clst) 
          
colourCount = length(unique(gath_df$K))
getPalette = colorRampPalette(brewer.pal(9, "Set3"))
p = ggplot(data=gath_df, aes(x=reorder(iid, desc(value)), y=value, fill=K)) + 
      geom_bar(stat="identity", width=1) +  
      #scale_fill_brewer(palette = "Set3") + 
      scale_fill_manual(values = getPalette(colourCount)) + 
      scale_y_continuous(expand=c(0, 0)) +
      scale_x_discrete(expand=c(0, 0)) +
      facet_grid(. ~ factor(clst, levels=pops), scales = "free", space="free", switch="both") + 
      theme_classic() +
      theme(panel.spacing = unit(0.2, "lines"), 
            strip.background = element_rect(colour="white", fill="white"),
            strip.text.x = element_text(colour = "black", angle = 90, hjust = 1.1), 
            strip.placement = "outside", 
            panel.grid.major = element_blank(), 
            panel.grid.minor = element_blank(),
            axis.text.x=element_blank(), 
            axis.ticks.x=element_blank()) + 
            ylab("") + 
            xlab("") +
            labs(fill="K") +
            theme(legend.position="bottom")
p
```
