---
title: "OutOfAfrica_3G09 [fixed shared factor]"
author: "Joseph Marcus"
date: "2020-05-18"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Here I visualize population structure with simulated data from the [OutOfAfrica_3G09](https://stdpopsim.readthedocs.io/en/latest/catalog.html#sec_catalog_homsap_models_outofafrica_3g09) scenario. See [Figure 2.](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000695) from Gutenkunst et al. 2009.

Below, I show a number of EBMF solutions and in each of them I don't display the first shared factor which is prefixed to the one-vector and scale the loadings by the prior variance. I only describe the loadings that remain after the shared factor.

# Imports

Import the required libraries and scripts:

```{r}
suppressMessages({
  library(lfa)
  library(flashier)
  library(drift.alpha)
  library(ggplot2)
  library(RColorBrewer)
  library(reshape2)
  library(tidyverse)
  library(alstructure)
  source("../code/structure_plot.R")
})
```

## Data

```{r}
data_path <- "../output/simulations/OutOfAfrica_3G09/rep1.txt"
G <- t(as.matrix(read.table(data_path, sep=" ")))
colnames(G) <- NULL
rownames(G) <- NULL
n <- nrow(G)
daf <- colSums(G) / (2 * n)
colors <- brewer.pal(8, "Set2")

# filter out too rare and too common SNPs
Y <- G[,((daf>=.05) & (daf <=.95))]
p <- ncol(Y)
print(n)
print(p)

# sub-population labels from stdpop
labs <- rep(c("YRI", "CEU", "HAN"), each=100)
```

we end up with 300 individuals and ~20000 SNPs. View fitted the sample covariance matrix:

```{r}
plot_cov((1.0 / p) * Y %*% t(Y) , as.is=T) + 
  scale_fill_viridis_c() + 
  labs(fill="Cov")
```

here is the sample correlation matrix:

```{r}
plot_cov(cov2cor((1.0 / p) * Y %*% t(Y)), as.is=T) + 
  scale_fill_viridis_c() + 
  labs(fill="Cov")
```

plot allele frequencies of Africa vs OOA populations:

```{r}
daf_afr <- colSums(G[1:100,]) / (2*100)
daf_ooa <- colSums(G[101:300,]) / (2*200)
qplot(daf_ooa, daf_afr, alpha=.1)
hist(daf_afr)
hist(daf_ooa)
```

# flash [greedy]

Run the `greedy` algorithm:

```{r}
ones <- matrix(1, nrow = n, ncol = 1)
ls.soln <- t(solve(crossprod(ones), crossprod(ones, Y)))
fl <- flash.init(Y) %>%
  flash.init.factors(EF = list(ones, ls.soln), 
                     prior.family=c(prior.bimodal(), prior.normal())) %>%
  flash.fix.loadings(kset = 1, mode = 1L) %>%
  flash.backfit() %>%
  flash.add.greedy(Kmax=6, prior.family=c(prior.bimodal(), prior.normal()))

sd <- unlist(lapply(fl$fitted.g[[2]], '[[', 3))
L <- fl$flash.fit$EF[[1]]
s2 <- fl$residuals.sd^2
LDsqrt <- L %*% diag(sd)
K <- ncol(LDsqrt)
plot_loadings(LDsqrt[,2:K], labs) + scale_color_brewer(palette="Set2")
```

view structure plot:

```{r}
create_structure_plot(L=LDsqrt[,2:K], labels=labs, colors=colors)
```

view fitted covariance matrix:

```{r, warning=FALSE, message=FALSE}
plot_cov(LDsqrt %*% t(LDsqrt) + s2*diag(n), as.is=T) + 
  scale_fill_viridis_c() +
  labs(fill="Cov") 
```

the `greedy` algorithm finds 3 population specific factors.

# flash [backfit]

Run `flash [backfit]` initializing from the greedy solution:

```{r}
flbf <- fl %>% 
  flash.backfit() %>% 
  flash.nullcheck(remove=TRUE)

sd <- unlist(lapply(flbf$fitted.g[[2]], '[[', 3))
s2 <- flbf$residuals.sd^2
L <- flbf$flash.fit$EF[[1]]
LDsqrt <- L %*% diag(sd)
K <- ncol(LDsqrt)
plot_loadings(LDsqrt[,2:K], labs) + scale_color_brewer(palette="Set2")
```

view structure plot:

```{r}
create_structure_plot(L=LDsqrt[,2:K], labels=labs, colors=colors)
```

view fitted covariance matrix:

```{r, warning=FALSE, message=FALSE}
plot_cov(LDsqrt %*% t(LDsqrt) + s2*diag(n), as.is=T) + 
  scale_fill_viridis_c() +
  labs(fill="Cov")
```

the `backfitting` algorithm represents the data with a sparser solution and finds a factor represented by YRI and a small loading from Han and 

# drift

Run `drift` initializing from the greedy solution:

```{r}
Kmax <- 10
EL_init <- matrix(runif(n*Kmax), nrow=n, ncol=Kmax)
EL_init[,1] <- 1
fl <- flash.init(Y) %>%
      flash.set.verbose(1L * TRUE)
ones <- matrix(1, nrow=nrow(Y), ncol=1)
ls.soln <- t(solve(crossprod(EL_init), crossprod(EL_init, Y)))      
fl <- fl %>%
      flash.init.factors(EF=list(EL_init, ls.soln),
                         prior.family=c(prior.bimodal(grid_size=100), 
                                        prior.normal())) %>%
      flash.fix.loadings(kset=1, mode=1L) %>%
      flash.backfit(maxiter=50, method="extrapolate")

init <- init_from_flash(fl)
plot(init$EL[,8])

dr <- drift(init, verbose=TRUE, extrapolate=TRUE, maxiter=2000, tol=1e-4)


#Kmax <- 6
#EL_init <- matrix(runif(n*Kmax), nrow=n, ncol=Kmax)
#EL_init[,1] <-1
#init <- init_from_EL(Y, EL_init,pm = )
#plot(init$EL[,1])

#Kmax <- 5
#EL_init <- matrix(runif(n*Kmax), nrow=n, ncol=Kmax)
#f <- nnfa(Y, EL_init, lower=0, upper=.1, tau=2)
#plot(f$L[,5])

#init <- init_from_EL(Y, f$L)

#Kmax <- 5
#init <- init_from_data(Y, Kmax=Kmax, incl.mean=TRUE)


sd <- sqrt(dr$prior_s2)
L <- dr$EL
LDsqrt <- L %*% diag(sd)
s2 <- dr$resid_s2
K <- ncol(LDsqrt)
plot_loadings(LDsqrt[,2:Kmax], labs) + scale_color_brewer(palette="Set2")
```

view structure plot:

```{r}
create_structure_plot(L=LDsqrt[,2:K], labels=labs, colors=colors)
```

view fitted covariance matrix:

```{r, warning=FALSE, message=FALSE}
plot_cov(LDsqrt %*% t(LDsqrt) + s2*diag(n), as.is=T) + 
  scale_fill_viridis_c() +
  labs(fill="Cov")
```

the `drift` algorithm converges quickly and maintains the greedy solution. Lets try adding a factor:

```{r}
dr2 <- drift(add_factor(dr), 
             miniter=2, 
             maxiter=1000, 
             tol=1e-5, 
             verbose=FALSE) 

sd <- sqrt(dr2$prior_s2)
L <- dr2$EL
LDsqrt <- L %*% diag(sd)
s2 <- dr2$resid_s2
K <- ncol(LDsqrt)
plot_loadings(LDsqrt[,2:K], labs) + scale_color_brewer(palette="Set2")
```

The additional factor does seem to capture much interesting structure. Lets now try intializing from a more shared representation of the data:

```{r}
n_samp_pop <- 100
EL <- cbind(c(rep(1, n_samp_pop), rep(1, n_samp_pop), rep(1, n_samp_pop)),
            c(rep(1, n_samp_pop), rep(0, n_samp_pop), rep(0, n_samp_pop)),
            c(rep(0, n_samp_pop), rep(1, n_samp_pop), rep(0, n_samp_pop)),
            c(rep(0, n_samp_pop), rep(0, n_samp_pop), rep(1, n_samp_pop)),
            c(rep(0, n_samp_pop), rep(1, n_samp_pop), rep(1, n_samp_pop))) 
EL <- EL + matrix(runif(5*n, 0,.05), nrow=n, ncol=5)
plot_loadings(EL, labs) 

EF <- t(solve(crossprod(EL), t(EL) %*% Y))
fl <- flash.init(Y, var.type = 0) %>%
      flash.init.factors(EF = list(EL, EF),
                         prior.family = c(prior.bimodal(), prior.normal())) %>%
      flash.fix.loadings(kset = 1, mode = 1L) %>%
      flash.backfit(maxiter=1) 

init <- init_from_flash(fl)
dr <- drift(init, miniter=2, 
            maxiter=1000, 
            tol=1e-3, 
            verbose=TRUE) 

sd <- sqrt(dr$prior_s2)
L <- dr$EL
LDsqrt <- L %*% diag(sd)
s2 <- dr$resid_s2
K <- ncol(LDsqrt)
plot_loadings(LDsqrt[,2:K], labs) + scale_color_brewer(palette="Set2")
```
