---
title: "OutOfAfrica_3G09"
author: "Joseph Marcus"
date: "2020-05-12"
output:
  workflowr::wflow_html:
    code_folding: hide
editor_options:
  chunk_output_type: console
---

Here I visualize population structure with simulated data from the [OutOfAfrica_3G09](https://stdpopsim.readthedocs.io/en/latest/catalog.html#sec_catalog_homsap_models_outofafrica_3g09) scenario. See [Figure 2.](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.1000695) from Gutenkunst et al. 2009.

# Imports

Import the required libraries and scripts:

```{r}
suppressMessages({
  library(lfa)
  library(flashier)
  library(drift.alpha)
  library(ggplot2)
  library(reshape2)
  library(tidyverse)
  library(alstructure)
})
```

# Data

```{r}
data_path <- "../output/simulations/OutOfAfrica_3G09/rep1.txt"
Y <- t(as.matrix(read.table(data_path, sep=" ")))
n <- nrow(Y)
maf <- colSums(Y) / (2 * n)

# filter out too rare and too common SNPs
Y <- Y[,((maf>=.05) & (maf <=.95))]
p <- ncol(Y)
Z <- scale(Y)
print(n)
print(p)

# sub-population labels from stdpop
labs <- rep(c("YRI", "CEU", "HAN"), each=40)
```

we end up with 120 individuals and ~30000 SNPs.

# PCA

Lets run `PCA` on the centered and scaled genotype matrix:

```{r}
svd_res <- lfa:::trunc.svd(Z, 5)
L_hat <- svd_res$u
plot_loadings(L_hat, labs)
```

plot the first two factors against each other:

```{r}
qplot(L_hat[,1], L_hat[,2], color=labs) + 
  xlab("PC1") + 
  ylab("PC2") + 
  theme_bw()
```

We certainly detect "clustered" population structure in the top PCs where PC1 represents Out of Africa and PC2 represents the next split between CEU and HAN.

# ALStructure

Run `ALStructure` with $K=3$:

```{r}
admix_res <- alstructure::alstructure(t(Y), d_hat=3)
Qhat <- t(admix_res$Q_hat)
plot_loadings(Qhat, labs)
```

the PSD model assigns three distinct clusters.

# flash [greedy]

Run the greedy algorithm:

```{r}
fl <- flash(Y, 
            greedy.Kmax=5, 
            prior.family=c(prior.bimodal(), prior.normal()))
plot_loadings(fl$flash.fit$EF[[1]], labs)
```

flash learns a shared factor between all the populations but where YRI has a lower loading then CEU and HAN and then it also learns population specific factors. 

# flash [backfit]

Run flash [backfit] initializing from the greedy solution:

```{r}
flbf <- fl %>% 
  flash.backfit() %>% 
  flash.nullcheck(remove=TRUE)
plot_loadings(flbf$flash.fit$EF[[1]], labs)
```

the backfitting algorithm represents the data with a sparser solution.

# drift

Run `drift` initializing from the greedy solution:

```{r}
init <- init_from_flash(fl)
dr <- drift(init, miniter=2, maxiter=500, tol=0.01, verbose=TRUE)
plot_loadings(dr$EL, labs)
```

drift seems to maintain the greedy solution.
