---
title: "Revisiting prior families for trees: Part V"
author: "Jason Willwerscheid"
date: "7/22/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 5, fig.height = 4)
```

```{r load.pkgs}
suppressMessages({
  library(flashier)
  library(drift.alpha)
  library(tidyverse)
})
```

Now I expand the tree to 8 populations. I include two admixed populations. Details are in the code.

```{r sim}
set.seed(666)

p <- 10000
resid_sd <- 0.1

# Define tree by mean branch length at each depth:
branch_means <- c(4, 2, 1, 0.5)
branch_sds <- c(0, 0.5, 0.3, 0.2)
depth <- length(branch_means)
npop_pure <- 2^(depth - 1)

# Define admixtures by admixture proportions:
admix_pops <- cbind(c(0, 0, 0, 0.4, 0.6, 0, 0, 0),
                   c(0, 0.15, 0.35, 0.5, 0, 0, 0, 0))
npop_admix <- ncol(admix_pops)

npop <- npop_pure + npop_admix

n <- sample(30:100, npop, replace = TRUE)
K <- 2^depth - 1

FF <- matrix(nrow = p, ncol = K)
k <- 1
for (d in 1:depth) {
  for (i in 1:(2^(d - 1))) {
    FF[, k] <- rnorm(p, sd = branch_means[d] + rnorm(1, sd = branch_sds[d]))
    k <- k + 1
  }
}

tree_mat <- matrix(0, nrow = npop_pure, ncol = K)
k <- 1
for (d in 1:depth) {
  size <- 2^(depth - d)
  for (i in 1:(2^(d - 1))) {
    tree_mat[((i - 1) * size + 1):(i * size), k] <- 1
    k <- k + 1
  }
}

pop_means <- FF %*% t(tree_mat)
pop_means <- cbind(pop_means, pop_means %*% admix_pops)

Y <- NULL
for (i in 1:npop) {
  Y <- rbind(Y, matrix(pop_means[, i], nrow = n[i], ncol = p, byrow = TRUE))
}
Y <- Y + rnorm(n * p, sd = resid_sd)

plot_dr <- function(dr) {
  sd <- sqrt(dr$prior_s2)
  L <- dr$EL
  LDsqrt <- L %*% diag(sd)
  K <- ncol(LDsqrt)
  plot_loadings(LDsqrt[,1:K], rep(letters[1:npop], n)) +
    scale_color_brewer(palette="Set3")
}

tree.fn = function(x, s, g_init, fix_g, output, ...) {
  if (is.null(g_init)) {
    g_init <- ashr::unimix(rep(1/3, 3), c(-1, 0, 1), c(-1, 0, 1))
  }

  return(flashier:::ebnm.nowarn(x = x,
                                s = s,
                                g_init = g_init,
                                fix_g = fix_g,
                                output = output,
                                prior_family = "ash",
                                prior = c(10, 1, 10),
                                ...))
}

prior.tree = function(...) {
  return(as.prior(sign = 0, ebnm.fn = function(x, s, g_init, fix_g, output) {
    tree.fn(x, s, g_init, fix_g, output, ...)
  }))
}

flextree.fn = function(a = 0, b = 1) {
  return(function(x, s, g_init, fix_g, output, ...) {
  if (is.null(g_init)) {
    g_init <- ashr::unimix(rep(1/5, 5), c(-1, 0, 1, -b, a), c(-1, 0, 1, -a, b))
  }

  return(flashier:::ebnm.nowarn(x = x,
                                s = s,
                                g_init = g_init,
                                fix_g = fix_g,
                                output = output,
                                prior_family = "ash",
                                prior = c(10, 10, 10, 1, 1),
                                ...))
  })
}

prior.flextree = function(...) {
  return(as.prior(sign = 0, ebnm.fn = function(x, s, g_init, fix_g, output) {
    flextree.fn(x, s, g_init, fix_g, output, ...)
  }))
}

init.mean.factor <- function(resids, zero.idx) {
  u <- matrix(1, nrow = nrow(resids), ncol = 1)
  u[zero.idx, 1] <- 0
  v <- t(solve(crossprod(u), crossprod(u, resids)))
  return(list(u, v))
}

# init.split.factor <- function(resids, zero.idx) {
#   svd.res <- svd(resids, nu = 1, nv = 1)
#   u <- svd.res$u
#   u[zero.idx] <- 0
#   u <- matrix(sign(u), ncol = 1)
#   v <- t(solve(crossprod(u), crossprod(u, resids)))
#   return(list(u, v))
# }
```

As long as the branch lengths are different enough one from the other, the simple method from the [previous analysis](pm1_priors4.html) works well enough. Here I progressively relax the priors in two stages to encourage the loadings to stay at -1, 0, and 1.

Initialization:

```{r fl}
fl <- flash.init(Y) %>%
  flash.set.verbose(0) %>%
  flash.init.factors(EF = init.mean.factor(Y, NULL), 
                     prior.family = c(prior.tree(), prior.normal())) %>%
  flash.fix.loadings(kset = 1, mode = 1L) %>%
  flash.backfit() %>%
  flash.add.greedy(Kmax = npop_pure - 1, 
                   prior.family = c(prior.tree(), prior.normal())) %>%
  flash.backfit()

plot_dr(init_from_flash(fl))
```

Partial relaxation:

```{r fl2}
fl2 <- fl

# Partial relaxation.
for (k in 1:fl2$n.factors) {
  fl2$flash.fit$ebnm.fn[[k]][[1]] <- flextree.fn(a = 0.2, b = 0.8)
}
fl2 <- fl2 %>% flash.backfit(warmstart = FALSE) %>% flash.backfit()

plot_dr(init_from_flash(fl2))
```

Full relaxation:

```{r fl3}
fl3 <- fl2

# Full relaxation.
for (k in 1:fl2$n.factors) {
  fl3$flash.fit$ebnm.fn[[k]][[1]] <- flextree.fn()
}
fl3 <- fl3 %>% flash.backfit(warmstart = FALSE) %>% flash.backfit()

plot_dr(init_from_flash(fl3))
```

The "fully relaxed" fit is a bit loose. For example, it appears from Factor 2 that Populations D and E are slightly admixed, but this is not the case. I suspect that the method will have a difficult time reliably finding small admixture proportions.
