---
title: "Revisiting prior families for trees: Part IX"
author: "Jason Willwerscheid"
date: "7/22/2020"
output:
  workflowr::wflow_html:
    code_folding: hide
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      fig.width = 5, fig.height = 4)
```

```{r load.pkgs}
suppressMessages({
  library(flashier)
  library(drift.alpha)
  library(tidyverse)
})
```

I return to the simplest scenario (a balanced tree with four populations of equal sizes) to see whether we can get the tree using a less contrived family of priors than the three-pointmass family.

```{r sim}
sim_tree <- function(n_range, 
                     p = 10000, 
                     branch_means, 
                     branch_sds, 
                     resid_sd = 0.1,
                     admix_pops = NULL, 
                     outgroup = FALSE, 
                     seed = 666) {
  set.seed(666)
  
  depth <- length(branch_means)
  npop_pure <- 2^(depth - 1)

  if (is.null(admix_pops)) {
    admix_pops <- matrix(nrow = 0, ncol = 0)
  }
  npop_admix <- ncol(admix_pops)
  
  npop <- npop_pure + npop_admix + outgroup
  
  if (length(n_range) == 1) {
    n <- rep(n_range, npop)
  } else {
    n <- sample(30:100, npop, replace = TRUE)
  }
  K <- 2^depth - 1
  
  FF <- matrix(nrow = p, ncol = K)
  k <- 1
  for (d in 1:depth) {
    for (i in 1:(2^(d - 1))) {
      FF[, k] <- rnorm(p, sd = branch_means[d] + rnorm(1, sd = branch_sds[d]))
      k <- k + 1
    }
  }
  
  tree_mat <- matrix(0, nrow = npop_pure, ncol = K)
  k <- 1
  for (d in 1:depth) {
    size <- 2^(depth - d)
    for (i in 1:(2^(d - 1))) {
      tree_mat[((i - 1) * size + 1):(i * size), k] <- 1
      k <- k + 1
    }
  }
  
  pop_means <- FF %*% t(tree_mat)
  if (npop_admix > 0) {
    pop_means <- cbind(pop_means, pop_means %*% admix_pops)
  }
  if (outgroup) {
    pop_means <- cbind(pop_means, rnorm(p, mean = 0, sd = sqrt(sum(branch_sds^2))))
  }
  
  Y <- NULL
  for (i in 1:npop) {
    Y <- rbind(Y, matrix(pop_means[, i], nrow = n[i], ncol = p, byrow = TRUE))
  }
  Y <- Y + rnorm(sum(n) * p, sd = resid_sd)
  
  plot_fl <- function(fl) {
    dr <- init_from_flash(fl)
    sd <- sqrt(dr$prior_s2)
    L <- dr$EL
    LDsqrt <- L %*% diag(sd)
    K <- ncol(LDsqrt)
    plot_loadings(LDsqrt[,1:K], rep(letters[1:npop], n)) +
      scale_color_brewer(palette="Set3")
  }
  
  return(list(Y = Y, plot_fn = plot_fl))
}
  
init.mean.factor <- function(resids, zero.idx) {
  u <- matrix(1, nrow = nrow(resids), ncol = 1)
  u[zero.idx, 1] <- 0
  v <- t(solve(crossprod(u), crossprod(u, resids)))
  return(list(u, v))
}
```

Normal mixture priors fail to find the correct rotation:

```{r normal_mix}
balanced_4pop <- sim_tree(n_range = 50,
                          p = 10000,
                          branch_means = rep(1, 3),
                          branch_sds = rep(0, 3),
                          resid_sd = 0.1)

fl_nsm <- flash.init(balanced_4pop$Y) %>%
  flash.set.verbose(0) %>%
  flash.add.greedy(Kmax = 4, 
                   prior.family = c(prior.normal.scale.mix(), prior.normal())) %>%
  flash.backfit(tol = 1e-4)

balanced_4pop$plot_fn(fl_nsm)
```

Unimodal priors can't find it either:

```{r unimix}
fl_uni <- flash.init(balanced_4pop$Y) %>%
  flash.set.verbose(0) %>%
  flash.add.greedy(Kmax = 4, 
                   prior.family = c(prior.unimodal(), prior.normal())) %>%
  flash.backfit(tol = 1e-4, maxiter = 1000, verbose.lvl = 0)

balanced_4pop$plot_fn(fl_uni)
```

Even varimax rotations won't find it:

```{r varimax}
add_split <- function(fl, Y, eigen.thresh = 0.5) {
  svd.res <- svd(Y - fitted(fl), nu = 10, nv = 10)
  n.eigen <- sum(svd.res$d > eigen.thresh * svd.res$d[1])
  if (n.eigen > 1) {
    varimax.res <- varimax(svd.res$u[, 1:n.eigen], normalize = FALSE)
    EL <- varimax.res$loadings[, 1, drop = FALSE]
    EF <- (svd.res$v[, 1:n.eigen] %*% t(solve(varimax.res$rotmat)))[, 1, drop = FALSE]
  } else {
    EL <- svd.res$u[, 1, drop = FALSE]
    EF <- svd.res$v[, 1, drop = FALSE]
  }
  fl.new <- fl %>%
    flash.init.factors(EF = list(EL, EF),
                       prior.family = c(prior.normal.scale.mix(), prior.normal())) %>%
    flash.backfit(kset = fl$n.factors + 1)
  return(fl.new)
}

fl_varimax <- flash.init(balanced_4pop$Y) %>%
  flash.set.verbose(0) %>%
  flash.add.greedy(Kmax = 2, 
                   prior.family = c(prior.normal.scale.mix(), prior.normal()))
for (i in 1:2) {
  fl_varimax <- add_split(fl_varimax, balanced_4pop$Y)
}
fl_varimax <- flash.backfit(fl_varimax, tol = 1e-4)

balanced_4pop$plot_fn(fl_varimax)
```

Point-Laplace priors, however, *do* find the tree!

```{r pl}
fl_pl <- flash.init(balanced_4pop$Y) %>%
  flash.set.verbose(0) %>%
  flash.add.greedy(Kmax = 4, 
                   prior.family = c(prior.point.laplace(), prior.normal())) %>%
  flash.backfit(tol = 1e-4, verbose.lvl = 0)

balanced_4pop$plot_fn(fl_pl)
```

