---
title: "Simplest Sree Simulation (bimodal prior)"
author: "jhmarcus"
date: "2019-05-21"
output: workflowr::wflow_html
---

In this analysis I simulate data from an extremely simple two population tree i.e. just a single split. See [code/sim.R](https://github.com/jhmarcus/drift-workflow/tree/master/code/sim.R) for simulation details.

# Import

Here I import the some required packages:

```{r imports, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(ashr)
library(flashier)
source("../code/viz.R")
source("../code/sim.R")
```

# Functions

```{r}
plot_flash_loadings = function(flash_fit, n_per_pop){

  l_df = as.data.frame(flash_fit$fit$EF[[1]])
  colnames(l_df) = 1:ncol(l_df)
  l_df$ID = 1:nrow(l_df)
  l_df$pop = c(rep("Pop1", n_per_pop), rep("Pop2", n_per_pop))
  
  gath_l_df = l_df %>% gather(K, value, -ID, -pop) 

  p1 = ggplot(gath_l_df, aes(x=ID, y=value, color=pop)) + 
       geom_point() +
       facet_wrap(K~., scale="free") +
       theme_bw() 
  
  p2 = structure_plot(gath_l_df %>% filter(K!=1), 
                      colset="Set3", 
                      facet_grp="pop", 
                      facet_levels=paste0("Pop", 1:2),
                      keep_leg=TRUE,
                      fact_type="nonnegative") 
  
  return(list(p1=p1, p2=p2))
  
}
```

# Simulate

Here I simulate a simple tree found [here](simpler_tree_simulation.html) with the below parameter settings:

```{r}
set.seed(123)

# generate data matrix
n_per_pop = 20
sigma_e = .1
p = 10000
sim_res = simplest_tree_simulation(n_per_pop, p, sigma_e)
Y = sim_res$Y

# plot the truth
l_df = as.data.frame(sim_res$L)
colnames(l_df) = 1:ncol(l_df)
l_df$ID = 1:nrow(l_df)
l_df$pop = c(rep("Pop1", n_per_pop), rep("Pop2", n_per_pop))
gath_l_df = l_df %>% gather(K, value, -ID, -pop) 
p = ggplot(gath_l_df, aes(x=ID, y=value, color=pop)) + 
    geom_point() +
    facet_wrap(K~., scale="free") +
    theme_bw() 
p
```

# Bimodal prior

Here I specify the grid of the prior

```{r}
# bimodal g prior list used throughout
m = 40
b = seq(1.0, 0.0, length=m)
a = seq(0.0, 1.0, length=m)
bimodal_g = ashr:::unimix(rep(0, 2*m), c(rep(0, m), b), c(a, rep(1, m)))
bimodal_g
```

# Bimodal (Greedy)

```{r}
flash_fit_g = flashier::flashier(Y, 
                                 greedy.Kmax=10,
                                 prior.type=c("nonnegative", "point.normal"),
                                 ash.param=list(fixg=FALSE, g=bimodal_g),
                                 ebnm.param=list(fix_pi0=TRUE, g=list(pi0=0.0)),
                                 var.type=0,
                                 fixed.factors=c(flashier:::ones.factor(n = 1)),
                                 backfit="none")

# plots
Lhat = flash_fit_g$fit$EF[[1]]
p_res = plot_flash_loadings(flash_fit_g, n_per_pop)
print(p_res$p1)
print(p_res$p2)
print(plot_covmat(Lhat %*% t(Lhat)))

# print summary of the fit
print(paste0("objective = ", flash_fit_g$objective))
print(paste0("estimated residual sd = ", sqrt(1 / flash_fit_g$fit$tau)))
print("estimated prior variances (factors) = ")
print(sapply(flash_fit_g$fit$g, function(x){1 / x[[2]]$a}))
```

# Bimodal (backfit)

```{r}
flash_fit_bf = flashier::flashier(Y, 
                                 greedy.Kmax=10,
                                 flash.init=flash_fit_g,
                                 prior.type=c("nonnegative", "normal.mixture"),
                                 ash.param=list(fixg=FALSE, g=bimodal_g),
                                 #ebnm.param=list(fix_pi0=TRUE, g=list(pi0=0.0)),
                                 var.type=0,
                                 backfit="final",
                                 backfit.order="random")

# plots
Lhat = flash_fit_bf$fit$EF[[1]]
p_res = plot_flash_loadings(flash_fit_bf, n_per_pop)
print(p_res$p1)
print(p_res$p2)
print(plot_covmat(Lhat %*% t(Lhat)))

# print summary of the fit
print(paste0("objective = ", flash_fit_bf$objective))
print(paste0("estimated residual sd = ", sqrt(1 / flash_fit_bf$fit$tau)))
print("estimated prior variances (factors) = ")
print(sapply(flash_fit_bf$fit$g, function(x){1 / x[[2]]$a}))
```